---
title: "Combined analysis of Chip and RNA-seq"
author: "Angelo Limeta"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
    theme: "journal"
editor_options:
  chunk_output_type: inline
always_allow_html: true
---

## Load Libraries
```{r}
library(tidyverse)

select = dplyr::select
```

## Load data
```{r}
tbl_meta <- readRDS(file = "../../results/processed_data/tbl_meta.rds")
tbl_rna <- readRDS(file = "../../results/processed_data/tbl_rna_cpm_with_meta.rds")
top_enhancers <- readRDS(file = "../../results/processed_data/top_enhancers.rds")

head(tbl_meta)
head(tbl_rna)
head(top_enhancers)
```

## Plot enhancer controlled gene expression

In this final quick analysis, I will simply check if the upregulation/downregulation of the predicted enhancer peaks correspond to a change in expression levels of their corrresponding gene.

```{r}
# Extract info from enhancer data
enhancer_ids <- top_enhancers$SYMBOL

tbl_rna_enh <- 
  tbl_rna %>% 
  filter(symbol %in% enhancer_ids)

tbl_rna_enh
```

Only one enhancer-gene is detected in the rna-seq data (PARD6B). Let's check its expression.
```{r}
p_PARD6B <- 
  tbl_rna_enh %>%
  ggplot(aes(x = condition,y = cpm, color = condition)) +
  geom_point() +
  theme_bw() +
  theme(aspect.ratio = 1,
        legend.position = "none") +
  scale_color_manual(values = c("grey60","darkred")) +
  labs(title = "RNA expression of PARD6B",
       x = "",
       y = "CPM")

p_PARD6B

ggsave(p_PARD6B,filename = "../../results/figures/combined_PARD6B_expression_rna.pdf")
  
```

In the Chip-data, this enhancer peak is downregulated in the KD condition, and this is also the case for the expression in the RNA-seq samples.

**PARD6B** (*Par-6 Family Cell Polarity Regulator Beta*):
Involved in maintaining cell polarity, which is crucial for proper cell adhesion and migration Downregulation could lead to disruption in cell adhesion, which aligns with reduced metastatic behavior. If cells are less able to maintain their polarity, they may have impaired motility, reducing their ability to spread.